---
title: "Whole Exome Sequencing Analysis Report"
subtitle: "SNV and INDEL Variant Prioritisation in IEI Candidate Genes"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
editor_options:
  chunk_output_type: inline
params:
  sample_id: "SAMPLE_1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "")

library(tidyverse)
library(knitr)
library(readxl)
library(vcfR)
library(kableExtra)
library(patchwork)
library(ggrepel)
library(svglite)
library(ragg)
library(showtext)
showtext_auto()

# ============================================================================
# NATURE-STYLE THEME (unified across all figures)
# ============================================================================
nature_theme <- theme_classic(base_size = 7, base_family = "Arial") +
  theme(
    axis.line        = element_line(linewidth = 0.35, colour = "black"),
    axis.ticks       = element_line(linewidth = 0.35, colour = "black"),
    axis.text        = element_text(size = 6.5, colour = "black"),
    axis.title       = element_text(size = 7),
    legend.title     = element_text(size = 6.5),
    legend.text      = element_text(size = 6),
    legend.position  = "bottom",
    legend.key.size  = unit(3, "mm"),
    strip.text       = element_text(size = 7, face = "bold"),
    plot.title       = element_text(size = 8, face = "bold", hjust = 0.5,
                                    margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 6.5, hjust = 0.5,
                                    margin = margin(b = 3)),
    plot.margin      = margin(4, 4, 2, 2),
    panel.grid       = element_blank(),
    panel.border     = element_blank()
  )

# Unified colour palette (low-saturation, Nature-compatible)
pal_primary   <- "#4472C4"   # blue — main data
pal_secondary <- "#8B6CAD"   # purple — secondary categories
pal_accent    <- "#ED7D31"   # orange — highlights / damaging variants
pal_pass      <- "#4D9E5D"   # green — QC PASS
pal_fail      <- "#D64545"   # red — QC FAIL
pal_neutral   <- "#9E9E9E"   # grey — reference lines
pal_bg        <- "#F5F5F5"   # light grey — background elements

# ============================================================================
# PUBLICATION FIGURE EXPORT
# ============================================================================
save_pub_r <- function(plot, filename, width_mm = 120, height_mm = 85, dpi = 600) {
  w <- width_mm / 25.4
  h <- height_mm / 25.4
  svglite::svglite(paste0(filename, ".svg"), width = w, height = h,
                   bg = "white", fix_text_size = FALSE)
  print(plot)
  dev.off()
  grDevices::cairo_pdf(paste0(filename, ".pdf"), width = w, height = h,
                       family = "Arial", bg = "white")
  print(plot)
  dev.off()
  ragg::agg_tiff(paste0(filename, ".tiff"), width = w, height = h,
                 units = "in", res = dpi, bg = "white")
  print(plot)
  dev.off()
}

# ============================================================================
# PATHS (parameterised by sample_id)
# ============================================================================
sample_id <- params$sample_id
sample_id_display <- gsub("_", "\\\\_", sample_id)  # escape underscore for LaTeX
input_dir <- "/path/to/your/R_analysis/directory"
out_dir   <- file.path(input_dir, "Marvis_output")
fig_dir   <- file.path(input_dir, "Figures", sample_id)
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

snp_csv    <- file.path(input_dir, paste0(sample_id, "_snp.hg38_multianno.csv"))
indel_csv  <- file.path(input_dir, paste0(sample_id, "_indel.hg38_multianno.csv"))
snp_vcf    <- file.path(input_dir, paste0(sample_id, "_snp.vcf"))
iei_file   <- file.path(input_dir, "IUIS-IEI-list-for-web-site-July-2024V2.xlsx")
const_file <- file.path(input_dir, "gnomad.v4.1.constraint_metrics.tsv")
```

\newpage

# Data Loading and Preprocessing

```{r load-data}
snp_multianno   <- readr::read_csv(snp_csv, col_names = TRUE, comment = "#",
                                    show_col_types = FALSE)
indel_multianno <- readr::read_csv(indel_csv, col_names = TRUE, comment = "#",
                                    show_col_types = FALSE)

# Load IEI gene panel
iei_data  <- read_excel(iei_file, sheet = 1)
iei_genes <- na.omit(unique(as.character(iei_data[["Genetic defect"]])))

# Build IEI inheritance lookup: Gene -> Inheritance (keep first match per gene)
# NOTE: distinct(.keep_all = FALSE) drops all non-grouping cols — use TRUE instead
iei_inheritance <- iei_data %>%
  rename(Gene = `Genetic defect`) %>%
  filter(!is.na(Gene), !is.na(Inheritance)) %>%
  mutate(Gene = str_trim(as.character(Gene)),
         Inheritance = str_trim(as.character(Inheritance))) %>%
  select(Gene, Inheritance) %>%
  distinct(Gene, .keep_all = TRUE)

# Load gnomAD constraint metrics (canonical transcripts only)
constraint_data <- read_tsv(const_file, comment = "#", show_col_types = FALSE)
constraint_lookup <- constraint_data %>%
  filter(canonical == TRUE | canonical == "true") %>%
  select(gene, lof.pLI, lof.oe_ci.upper) %>%
  rename(Gene = gene, pLI = lof.pLI, LOEUF = lof.oe_ci.upper) %>%
  distinct(Gene, .keep_all = TRUE)

if (nrow(constraint_lookup) == 0) {
  constraint_lookup <- constraint_data %>%
    select(gene, lof.pLI, lof.oe_ci.upper) %>%
    rename(Gene = gene, pLI = lof.pLI, LOEUF = lof.oe_ci.upper) %>%
    distinct(Gene, .keep_all = TRUE)
}

cat(sprintf("Sample: %s | IEI genes: %d | Constraint genes: %d\n",
            sample_id, length(iei_genes), nrow(constraint_lookup)))
```

\newpage

# Coverage Overview

```{r coverage-overview, fig.height=3.5, fig.width=7}
# ---- Helper: covered genes and transcripts ----
get_covered_genes <- function(df) {
  df %>%
    filter(Func.refGene %in% c("exonic", "exonic;splicing", "splicing")) %>%
    filter(grepl("NM_", AAChange.refGene)) %>%
    separate_rows(Gene.refGene, sep = ",") %>%
    filter(!grepl("NONE", Gene.refGene)) %>%
    mutate(Gene = str_trim(Gene.refGene)) %>%
    filter(nchar(Gene) > 0, Gene != ".") %>%
    pull(Gene) %>% unique()
}

get_covered_transcripts <- function(df) {
  df %>%
    filter(Func.refGene %in% c("exonic", "exonic;splicing", "splicing")) %>%
    filter(grepl("NM_", AAChange.refGene)) %>%
    pull(AAChange.refGene) %>% str_split(",") %>% unlist() %>%
    str_extract("NM_[0-9]+\\.?[0-9]*") %>% na.omit() %>% unique()
}

snp_genes         <- get_covered_genes(snp_multianno)
indel_genes       <- get_covered_genes(indel_multianno)
snp_transcripts   <- get_covered_transcripts(snp_multianno)
indel_transcripts <- get_covered_transcripts(indel_multianno)
all_genes         <- union(snp_genes, indel_genes)
all_transcripts   <- union(snp_transcripts, indel_transcripts)
total_human_pcg   <- 20076
pct_covered <- round(length(all_genes) / total_human_pcg * 100, 2)

# ---- Coverage summary table ----
coverage_summary <- tibble(
  Metric = c(
    "Unique genes (SNV)", "Unique genes (INDEL)", "Unique genes (combined)",
    "Unique transcripts (SNV)", "Unique transcripts (INDEL)", "Unique transcripts (combined)",
    "Human protein-coding genes (RefSeq hg38)", "% Human protein-coding genes detected"
  ),
  Value = c(
    length(snp_genes), length(indel_genes), length(all_genes),
    length(snp_transcripts), length(indel_transcripts), length(all_transcripts),
    total_human_pcg, paste0(pct_covered, "%")
  )
)

knitr::kable(coverage_summary,
  caption = "Gene and Transcript Coverage Summary (exonic/splicing variants)",
  align = c("l", "r"), booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")

# ---- Fig 1: Coverage bar ----
coverage_plot_df <- tibble(
  Category = factor(c("SNV genes", "INDEL genes", "Combined genes",
                       "SNV transcripts", "INDEL transcripts", "Combined transcripts"),
    levels = c("SNV genes", "INDEL genes", "Combined genes",
               "SNV transcripts", "INDEL transcripts", "Combined transcripts")),
  Type  = c(rep("Genes", 3), rep("Transcripts", 3)),
  Count = c(length(snp_genes), length(indel_genes), length(all_genes),
            length(snp_transcripts), length(indel_transcripts), length(all_transcripts))
)

p_coverage <- ggplot(coverage_plot_df, aes(x = Category, y = Count, fill = Type)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = scales::comma(Count)), vjust = -0.4, size = 2.2) +
  scale_fill_manual(values = c("Genes" = pal_primary, "Transcripts" = pal_secondary)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)), labels = scales::comma) +
  labs(title = "WES Coverage Overview",
       subtitle = paste0(sample_id_display, " — ", pct_covered, "% of human protein-coding genes detected"),
       x = NULL, y = "Count", fill = NULL) +
  nature_theme + theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 6))

p_coverage
save_pub_r(p_coverage, file.path(fig_dir, paste0(sample_id, "_Fig1_coverage")), 140, 85)

# ---- SNV vs INDEL gene overlap ----
snv_only  <- length(setdiff(snp_genes, indel_genes))
indel_only <- length(setdiff(indel_genes, snp_genes))
both <- length(intersect(snp_genes, indel_genes))

overlap_df <- tibble(
  Category = c("SNV only", "Both", "INDEL only"),
  Count = c(snv_only, both, indel_only)
)

p_overlap <- ggplot(overlap_df, aes(x = factor(Category,
      levels = c("SNV only", "Both", "INDEL only")), y = Count, fill = Category)) +
  geom_col(width = 0.55, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = Count), vjust = -0.4, size = 2.5) +
  scale_fill_manual(values = c("SNV only" = pal_primary,
                                "Both" = pal_accent,
                                "INDEL only" = pal_secondary)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Gene Overlap: SNV vs INDEL",
       subtitle = paste("Shared:", both, "genes"),
       x = NULL, y = "Number of genes") +
  nature_theme + theme(legend.position = "none")

p_overlap
save_pub_r(p_overlap, file.path(fig_dir, paste0(sample_id, "_Fig2_overlap")), 85, 75)
```

\newpage

# Functional Annotation Summary

## SNV Functional Annotations

```{r snv-functional, fig.height=9, fig.width=7}
# ---- (a) Functional category distribution ----
func_counts_snp <- snp_multianno %>%
  count(Func.refGene, sort = TRUE) %>%
  rename(Function = Func.refGene, Count = n)

p_func <- ggplot(func_counts_snp, aes(x = reorder(Function, Count), y = Count)) +
  geom_col(fill = pal_primary, width = 0.7, colour = "white", linewidth = 0.2) +
  geom_text(aes(label = Count), hjust = -0.2, size = 2.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)), labels = scales::comma) +
  labs(title = "SNV Functional Categories", x = NULL, y = "Count") +
  nature_theme

# ---- (b) Exonic variant types ----
variant_summary_snp <- snp_multianno %>%
  filter(Func.refGene %in% c("exonic", "exonic;splicing")) %>%
  count(ExonicFunc.refGene) %>%
  arrange(desc(n)) %>%
  filter(ExonicFunc.refGene != "." & !is.na(ExonicFunc.refGene))

p_vartype <- ggplot(variant_summary_snp,
  aes(x = reorder(ExonicFunc.refGene, n), y = n)) +
  geom_col(fill = pal_secondary, width = 0.7, colour = "white", linewidth = 0.2) +
  geom_text(aes(label = n), hjust = -0.2, size = 2.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Exonic SNV Variant Types", x = NULL, y = "Count") +
  nature_theme

# ---- (c) Top genes by variant count ----
top_genes_snp <- snp_multianno %>%
  filter(!grepl("NONE", Gene.refGene), Gene.refGene != ".") %>%
  count(Gene.refGene, sort = TRUE) %>%
  slice_head(n = 10) %>%
  rename(Gene = Gene.refGene, Count = n) %>%
  mutate(Gene_label = str_trunc(Gene, 14))

p_topgenes <- ggplot(top_genes_snp, aes(x = reorder(Gene_label, Count), y = Count)) +
  geom_col(fill = pal_primary, width = 0.7, colour = "white", linewidth = 0.2) +
  geom_text(aes(label = Count), hjust = -0.2, size = 2.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.14))) +
  labs(title = "Top Genes with Most SNV Variants", x = NULL, y = "Count") +
  nature_theme + theme(axis.text.y = element_text(size = 5.5))

# ---- (d) Rare variant genes ----
rare_variants_snp <- snp_multianno %>%
  mutate(gnomad_af = suppressWarnings(as.numeric(na_if(gnomad41_exome_AF, ".")))) %>%
  filter(!is.na(gnomad_af) & gnomad_af < 0.01) %>%
  filter(!grepl("NONE", Gene.refGene), Gene.refGene != ".") %>%
  count(Gene.refGene, sort = TRUE) %>%
  slice_head(n = 10) %>%
  rename(Gene = Gene.refGene, Count = n) %>%
  mutate(Gene_label = str_trunc(Gene, 14))

p_rare <- ggplot(rare_variants_snp, aes(x = reorder(Gene_label, Count), y = Count)) +
  geom_col(fill = pal_accent, width = 0.7, colour = "white", linewidth = 0.2) +
  geom_text(aes(label = Count), hjust = -0.2, size = 2.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Top Genes with Rare SNVs (gnomAD AF < 1%)",
       x = NULL, y = "Count") +
  nature_theme + theme(axis.text.y = element_text(size = 5.5))

# ---- Assemble 4-panel layout ----
p_snv_composite <- (p_func | p_vartype) / (p_topgenes | p_rare) +
  plot_annotation(
    title = paste0(sample_id_display, " — SNV Functional Annotation"),
    tag_levels = "a"
  ) & theme(plot.tag = element_text(size = 8, face = "bold"))

p_snv_composite
save_pub_r(p_snv_composite, file.path(fig_dir, paste0(sample_id, "_Fig3_SNV_annotation")), 183, 180)
```

\newpage

## INDEL Functional Annotations

```{r indel-functional, fig.height=9, fig.width=7}
# ---- (a) Functional categories ----
func_counts_indel <- indel_multianno %>%
  count(Func.refGene, sort = TRUE) %>%
  rename(Function = Func.refGene, Count = n)

p_ifunc <- ggplot(func_counts_indel, aes(x = reorder(Function, Count), y = Count)) +
  geom_col(fill = pal_secondary, width = 0.7, colour = "white", linewidth = 0.2) +
  geom_text(aes(label = Count), hjust = -0.2, size = 2.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "INDEL Functional Categories", x = NULL, y = "Count") +
  nature_theme

# ---- (b) Allele frequency (guarded NA handling) ----
indel_af <- indel_multianno %>%
  mutate(gnomad_af = suppressWarnings(as.numeric(na_if(gnomad41_exome_AF, ".")))) %>%
  filter(!is.na(gnomad_af))

n_na_af <- sum(is.na(suppressWarnings(
  as.numeric(na_if(indel_multianno[["gnomad41_exome_AF"]], ".")))))

p_af <- ggplot(indel_af, aes(x = gnomad_af)) +
  geom_histogram(bins = 50, fill = pal_secondary, colour = "white", linewidth = 0.15) +
  geom_vline(xintercept = 0.01, linetype = "dashed", colour = pal_accent, linewidth = 0.5) +
  annotate("text", x = 0.015, y = Inf, label = "AF = 1%", colour = pal_accent,
           size = 2.2, hjust = 0, vjust = 2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(title = "INDEL Allele Frequency (gnomAD v4.1)",
       subtitle = paste(n_na_af, "variants without gnomAD annotation (retained)"),
       x = "Allele Frequency", y = "Count") +
  nature_theme

# ---- (c) Chromosome distribution (Nature-Optimized Fix) ----
chr_counts_indel <- indel_multianno %>%
  count(Chr, sort = TRUE) %>%
  filter(str_detect(Chr, "^chr([1-9]|1[0-9]|2[0-2]|[XYM])$")) %>%
  mutate(Chr = factor(Chr, levels = c(paste0("chr", 1:22), "chrX", "chrY", "chrM"))) %>%
  filter(!is.na(Chr))

p_chr <- ggplot(chr_counts_indel, aes(x = Chr, y = n)) +
  geom_col(fill = pal_primary, width = 0.6, colour = "white", linewidth = 0.15) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.4, size = 2.5, fontface = "bold", color = "#333333") +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = scales::comma
  ) +
  labs(
    title = "Chromosome Distribution of INDELs",
    x = "Chromosome",
    y = "Number of INDELs"
  ) +
  nature_theme +
  theme(
    axis.text.x = element_text(angle = 55, hjust = 1, size = 8, color = "#222222"),
    axis.title.x = element_text(size = 10, margin = margin(t = 10)),
    plot.title = element_text(size = 8, face = "bold", hjust = 0.5, pad = 10)
  )

# ---- (d) Insertion vs Deletion ----
indel_types <- indel_multianno %>%
  mutate(IndelType = case_when(
    nchar(Ref) < nchar(Alt) ~ "Insertion",
    nchar(Ref) > nchar(Alt) ~ "Deletion",
    TRUE ~ "Other"
  )) %>%
  count(IndelType, sort = TRUE) %>%
  filter(IndelType != "Other")

p_type <- ggplot(indel_types, aes(x = IndelType, y = n, fill = IndelType)) +
  geom_col(width = 0.45, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = n), vjust = -0.4, size = 2.5) +
  scale_fill_manual(values = c("Insertion" = pal_accent, "Deletion" = pal_secondary)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "INDEL Types", x = NULL, y = "Count") +
  nature_theme + theme(legend.position = "none")

# ---- Assemble ----
p_indel_composite <- (p_ifunc | p_af) / (p_chr | p_type) +
  plot_annotation(
    title = paste0(sample_id_display, " — INDEL Functional Annotation"),
    tag_levels = "a"
  ) & theme(plot.tag = element_text(size = 8, face = "bold"))

p_indel_composite
save_pub_r(p_indel_composite,
  file.path(fig_dir, paste0(sample_id, "_Fig4_INDEL_annotation")), 183, 180)
```

\newpage

# Reference Data

```{r ref-data}
cat(sprintf("**IEI Gene Panel:** %d genes loaded from IUIS 2024 list.\n\n",
            length(iei_genes)))
cat(sprintf("**gnomAD Constraint Metrics:** %d genes (canonical transcripts only).\n",
            nrow(constraint_lookup)))
```

\newpage

# Variant Filtering and Prioritisation

## Filtering Strategy

**SNV filters** (sequential):
1. Exonic / exonic;splicing regions
2. IEI candidate gene list
3. Exclude synonymous SNVs
4. gnomAD v4.1 exome AF < 1% (NA retained)
5. In-silico: AlphaMissense (pathogenic/likely_pathogenic) | REVEL > 0.75 | CADD >= 20 | stopgain/stoploss/startloss
6. Annotate pLI + LOEUF (not used for filtering)

**INDEL filters** (sequential):
1. Exonic / exonic;splicing / splicing regions
2. IEI candidate gene list
3. Non-synonymous / frameshift / stop-altering types
4. gnomAD v4.1 exome AF < 1% (NA retained)
5. Frameshift/stopgain = unconditional pass; nonframeshift requires CADD >= 15
6. Annotate pLI + LOEUF

```{r snv-filtering}
pop_freq_threshold <- 0.01

filter_log <- tibble(Stage = character(), Variants = integer())

# ---- SNV Pipeline ----
total_snvs <- nrow(snp_multianno)
filter_log <- filter_log %>% add_row(Stage = "Total SNVs", Variants = total_snvs)

exonic_snvs <- snp_multianno %>%
  filter(Func.refGene %in% c("exonic", "exonic;splicing"))
filter_log <- filter_log %>% add_row(Stage = "Exonic", Variants = nrow(exonic_snvs))

iei_snvs <- exonic_snvs %>% filter(Gene.refGene %in% iei_genes)
filter_log <- filter_log %>%
  add_row(Stage = "In IEI genes", Variants = nrow(iei_snvs))

nonsyn_snvs <- iei_snvs %>%
  filter(ExonicFunc.refGene %in% c("nonsynonymous SNV", "stopgain", "stoploss",
                                    "startloss", "unknown"))
filter_log <- filter_log %>%
  add_row(Stage = "Non-synonymous", Variants = nrow(nonsyn_snvs))

rare_snvs <- nonsyn_snvs %>%
  mutate(gnomad_af = suppressWarnings(as.numeric(na_if(gnomad41_exome_AF, ".")))) %>%
  filter(is.na(gnomad_af) | gnomad_af < pop_freq_threshold)
filter_log <- filter_log %>%
  add_row(Stage = "gnomAD AF < 1%", Variants = nrow(rare_snvs))

damaging_snvs <- rare_snvs %>%
  mutate(
    revel_score = suppressWarnings(as.numeric(na_if(REVEL_score, "."))),
    cadd_score  = suppressWarnings(as.numeric(na_if(CADD_phred, ".")))
  ) %>%
  filter(
    ExonicFunc.refGene %in% c("stopgain", "stoploss", "startloss") |
      AlphaMissense_pred %in% c("pathogenic", "likely_pathogenic") |
      (!is.na(revel_score) & revel_score > 0.75) |
      (!is.na(cadd_score) & cadd_score >= 20)
  )
filter_log <- filter_log %>%
  add_row(Stage = "Pathogenicity filter", Variants = nrow(damaging_snvs))

final_variants <- damaging_snvs %>%
  left_join(constraint_lookup, by = c("Gene.refGene" = "Gene"))
filter_log <- filter_log %>%
  add_row(Stage = "Final candidate SNVs", Variants = nrow(final_variants))

write.csv(final_variants,
  file.path(out_dir, paste0(sample_id, "_candidate_SNVs.csv")), row.names = FALSE)

# ---- INDEL Pipeline ----
indel_log <- tibble(Stage = character(), Variants = integer())

total_indels <- nrow(indel_multianno)
indel_log <- indel_log %>% add_row(Stage = "Total INDELs", Variants = total_indels)

exonic_indels <- indel_multianno %>%
  filter(Func.refGene %in% c("exonic", "exonic;splicing", "splicing"))
indel_log <- indel_log %>%
  add_row(Stage = "Exonic/splicing", Variants = nrow(exonic_indels))

iei_indels <- exonic_indels %>% filter(Gene.refGene %in% iei_genes)
indel_log <- indel_log %>%
  add_row(Stage = "In IEI genes", Variants = nrow(iei_indels))

indel_priority <- c("frameshift deletion", "frameshift insertion",
  "nonframeshift deletion", "nonframeshift insertion",
  "stopgain", "stoploss", "startloss", "unknown")
nonsyn_indels <- iei_indels %>% filter(ExonicFunc.refGene %in% indel_priority)
indel_log <- indel_log %>%
  add_row(Stage = "Non-synonymous/fs", Variants = nrow(nonsyn_indels))

rare_indels <- nonsyn_indels %>%
  mutate(gnomad_af = suppressWarnings(as.numeric(na_if(gnomad41_exome_AF, ".")))) %>%
  filter(is.na(gnomad_af) | gnomad_af < pop_freq_threshold)
indel_log <- indel_log %>%
  add_row(Stage = "gnomAD AF < 1%", Variants = nrow(rare_indels))

frameshift_types <- c("frameshift deletion", "frameshift insertion",
                       "stopgain", "stoploss", "startloss")

damaging_indels <- rare_indels %>%
  mutate(
    cadd_score = suppressWarnings(as.numeric(na_if(CADD_phred, "."))),
    is_damaging = case_when(
      ExonicFunc.refGene %in% frameshift_types ~ TRUE,
      ExonicFunc.refGene %in% c("nonframeshift deletion", "nonframeshift insertion") &
        !is.na(cadd_score) & cadd_score >= 15 ~ TRUE,
      TRUE ~ FALSE
    )
  ) %>%
  filter(is_damaging)
indel_log <- indel_log %>%
  add_row(Stage = "Pathogenicity filter", Variants = nrow(damaging_indels))

final_indels <- damaging_indels %>%
  left_join(constraint_lookup, by = c("Gene.refGene" = "Gene"))
indel_log <- indel_log %>%
  add_row(Stage = "Final candidate INDELs", Variants = nrow(final_indels))

write.csv(final_indels,
  file.path(out_dir, paste0(sample_id, "_candidate_INDELs.csv")), row.names = FALSE)
```


```{r filtering-waterfall, fig.height=6, fig.width=9}
# ---- Fig: SNV Filtering Waterfall ----

filter_log <- filter_log %>%
  mutate(
    Stage = factor(Stage, levels = Stage),
    label_col = ifelse(
      Stage == "Final candidate SNVs",
      pal_accent,
      pal_primary
    )
  )

p_snv_funnel <- ggplot(filter_log, aes(x = Stage, y = Variants)) +
  geom_col(
    aes(fill = label_col),
    width = 0.65,
    colour = "white",
    linewidth = 0.3
  ) +
  geom_text(
    aes(label = scales::comma(Variants)),
    vjust = -0.4,
    size = 2.8,
    fontface = "bold"
  ) +
  scale_fill_identity() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = scales::comma
  ) +
  labs(
    title = paste0(sample_id_display, " — SNV Filtering Cascade"),
    subtitle = paste(
      "Final:",
      nrow(final_variants),
      "candidate SNVs from",
      scales::comma(total_snvs),
      "input variants"
    ),
    x = "Filtering Stage",
    y = "Number of Variants"
  ) +
  nature_theme +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1,
      size = 8.5,
      colour = "black",
      face = "bold"
    ),
    axis.ticks.x = element_line(colour = "black"),
    plot.title = element_text(
      size = 13,
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      hjust = 0.5,
      face = "italic",
      colour = "#444444"
    )
  )

p_snv_funnel

save_pub_r(
  p_snv_funnel,
  file.path(fig_dir, paste0(sample_id, "_Fig5_SNV_waterfall")),
  160,
  110
)


# ---- Fig: INDEL Filtering Waterfall ----

indel_log <- indel_log %>%
  mutate(
    Stage = factor(Stage, levels = Stage),
    label_col = ifelse(
      Stage == "Final candidate INDELs",
      pal_accent,
      pal_secondary
    )
  )

p_indel_funnel <- ggplot(indel_log, aes(x = Stage, y = Variants)) +
  geom_col(
    aes(fill = label_col),
    width = 0.65,
    colour = "white",
    linewidth = 0.3
  ) +
  geom_text(
    aes(label = scales::comma(Variants)),
    vjust = -0.4,
    size = 2.8,
    fontface = "bold"
  ) +
  scale_fill_identity() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15)),
    labels = scales::comma
  ) +
  labs(
    title = paste0(sample_id_display, " — INDEL Filtering Cascade"),
    subtitle = paste(
      "Final:",
      nrow(final_indels),
      "candidate INDELs from",
      scales::comma(total_indels),
      "input variants"
    ),
    x = "Filtering Stage",
    y = "Number of Variants"
  ) +
  nature_theme +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1,
      size = 8.5,
      colour = "black",
      face = "bold"
    ),
    axis.ticks.x = element_line(colour = "black"),
    plot.title = element_text(
      size = 13,
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      hjust = 0.5,
      face = "italic",
      colour = "#444444"
    )
  )

p_indel_funnel

save_pub_r(
  p_indel_funnel,
  file.path(fig_dir, paste0(sample_id, "_Fig6_INDEL_waterfall")),
  160,
  110
)
```



\newpage

```{r filtering-table, fig.height=6, fig.width=9}
# ---- SNV Filtering Funnel Summary Table ----
snv_summary <- tibble(
  `Sequential Filtering Stage` = filter_log$Stage,
  `Remaining SNV Count` = scales::comma(filter_log$Variants),
)

knitr::kable(snv_summary,
             caption = paste0(
    "Sequential Variant Filtering Cascade Summary for SNVs (",
    sample_id_display,
    ")"
  ),
             align = c("l", "r"),
             booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")


# ---- INDEL Filtering Funnel Summary Table ----
indel_summary <- tibble(
  `Sequential Filtering Stage` = indel_log$Stage,
  `Remaining INDEL Count` = scales::comma(indel_log$Variants),
)

knitr::kable(indel_summary,
             caption = paste0("Sequential Variant Filtering Cascade Summary for INDELs (",
    sample_id_display,
    ")"
  ),
             align = c("l", "r"),
             booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position")

```



```{r filtering-tables}
snv_display_cols <- intersect(c(
  "Chr", "Gene.refGene", "ExonicFunc.refGene",
  "gnomad41_exome_AF", "AlphaMissense_pred", "pLI", "LOEUF"
), names(final_variants))

knitr::kable(final_variants[, snv_display_cols],
  caption = "Final Candidate SNVs with Constraint Metrics",
  digits = 3, booktabs = TRUE,
  col.names = c("Chr", "Gene", "Exonic Function", "gnomAD AF",
                "AlphaMissense", "pLI", "LOEUF")) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                font_size = 7)

indel_display_cols <- intersect(c(
  "Chr", "Gene.refGene", "ExonicFunc.refGene",
  "gnomad41_exome_AF", "CADD_phred", "pLI", "LOEUF"
), names(final_indels))


knitr::kable(final_indels[, indel_display_cols],
  caption = "Final Candidate INDELs with Constraint Metrics",
  digits = 3, booktabs = TRUE,
  col.names = c("Chr", "Gene", "Exonic Function", "gnomAD AF",
                "CADD", "pLI", "LOEUF")) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                font_size = 7)
```

\newpage

# Genotype Reconstruction and Quality Control

```{r genotype-qc, fig.height=5, fig.width=8}
# ---- Genotype extraction from VCF ----
candidate_variants <- read_csv(
  file.path(out_dir, paste0(sample_id, "_candidate_SNVs.csv")),
  show_col_types = FALSE)

vcf_data <- read.vcfR(snp_vcf, verbose = FALSE)

vcf_lookup <- tibble(
  Chr   = getCHROM(vcf_data),
  Start = as.numeric(getPOS(vcf_data)),
  Ref   = getREF(vcf_data),
  Alt   = str_split(getALT(vcf_data), ",", simplify = TRUE)[, 1],
  QUAL  = as.numeric(getQUAL(vcf_data)),
  GT    = extract.gt(vcf_data, element = "GT")[, 1],
  DP    = as.numeric(extract.gt(vcf_data, element = "DP")[, 1]),
  AD    = extract.gt(vcf_data, element = "AD")[, 1]
) %>%
  mutate(
    AD_ref = suppressWarnings(as.numeric(str_split(AD, ",", simplify = TRUE)[, 1])),
    AD_alt = suppressWarnings(as.numeric(str_split(AD, ",", simplify = TRUE)[, 2])),
    AF_obs = case_when(
      !is.na(AD_ref) & !is.na(AD_alt) & (AD_ref + AD_alt) > 0 ~
        AD_alt / (AD_ref + AD_alt),
      TRUE ~ NA_real_),
    zygosity = case_when(
      GT %in% c("0/1", "0|1", "1/0", "1|0") ~ "Heterozygous",
      GT %in% c("1/1", "1|1") ~ "Homozygous",
      GT %in% c("0/0", "0|0") ~ "Reference",
      TRUE ~ "Unknown"
    )
  )

genotype_table <- candidate_variants %>%
  left_join(vcf_lookup %>% select(Chr, Start, Ref, Alt, GT, DP, QUAL, AF_obs, zygosity),
            by = c("Chr", "Start", "Ref", "Alt")) %>%
  mutate(
    variant_id = paste0(Chr, ":", Start, ":", Ref, ">", Alt),
    pass_depth = DP >= 10,
    pass_qual  = QUAL >= 20,
    QC_status  = if_else(pass_depth & pass_qual, "PASS", "FAIL"),
    QC_flag    = case_when(
      !pass_depth & !pass_qual ~ "LOW_DEPTH;LOW_QUAL",
      !pass_depth ~ "LOW_DEPTH",
      !pass_qual ~ "LOW_QUAL",
      TRUE ~ "PASS"
    )
  )

write_csv(genotype_table,
  file.path(out_dir, paste0(sample_id, "_genotype_with_qc.csv")))

# ---- Fig: QC scatter (DP vs Allele Fraction) ----
qc_plot_data <- genotype_table %>% filter(!is.na(DP) & !is.na(AF_obs))

if (nrow(qc_plot_data) > 0) {
  p_qc <- ggplot(qc_plot_data, aes(x = DP, y = AF_obs)) +
    geom_vline(xintercept = 10, linetype = "dashed", colour = pal_neutral,
               linewidth = 0.4) +
    geom_hline(yintercept = c(0.25, 0.75), linetype = "dashed",
               colour = pal_neutral, linewidth = 0.4) +
    geom_hline(yintercept = 0.5, linetype = "solid",
               colour = pal_primary, alpha = 0.3, linewidth = 0.4) +
    geom_point(aes(fill = QC_status), size = 3.5, shape = 21,
               colour = "grey30", stroke = 0.3, alpha = 0.85) +
    geom_text_repel(aes(label = variant_id), size = 2, max.overlaps = 15,
                    colour = "grey40") +
    annotate("rect", xmin = 10, xmax = Inf, ymin = 0.25, ymax = 0.75,
             fill = pal_pass, alpha = 0.05) +
    scale_fill_manual(values = c("PASS" = pal_pass, "FAIL" = pal_fail)) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.25, 0.5, 0.75, 1)) +
    labs(title = paste0(sample_id_display, " — Candidate SNV Quality Control"),
         subtitle = "Read depth vs. observed allele fraction | Dashed: het. balance bounds (0.25–0.75)",
         x = "Read Depth (DP)", y = "Alternate Allele Fraction",
         fill = "QC Status") +
    nature_theme


  save_pub_r(p_qc, file.path(fig_dir, paste0(sample_id, "_Fig7_QC")), 160, 110)
  p_qc
} else {
  cat("No candidate variants to plot QC.\n")
}

# ---- QC summary table ----
knitr::kable(genotype_table %>%
  select(variant_id, Gene = Gene.refGene, ExonicFunc = ExonicFunc.refGene,
         zygosity, DP, QUAL, AF_obs, QC_status, QC_flag),
  caption = "Candidate SNVs: Genotype and Quality Control",
  digits = 3, booktabs = TRUE,
  col.names = c("Variant", "Gene", "Exonic Function", "Zygosity",
                "DP", "QUAL", "AF", "QC", "QC Flag")) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                font_size = 7)
```

\newpage

# Constraint Metrics Overview

```{r constraint-scatter, fig.height=4.5, fig.width=7}
# ---- Fig 8: pLI vs LOEUF for candidate genes (SNV + INDEL combined) ----
# FIX: pool SNV and INDEL candidate genes together so the plot is never empty
# when one set has no constraint matches, and drop rows where both pLI and
# LOEUF are NA (which previously produced a blank but non-empty data frame).

constraint_plot_df <- bind_rows(
    final_variants %>% select(Gene.refGene, pLI, LOEUF),
    final_indels   %>% select(Gene.refGene, pLI, LOEUF)
  ) %>%
  distinct() %>%
  filter(!(is.na(pLI) & is.na(LOEUF))) %>%   # drop genes with NO constraint data
  mutate(
    constraint_class = case_when(
      !is.na(pLI)   & pLI   >= 0.9  ~ "LoF intolerant (pLI >= 0.9)",
      !is.na(LOEUF) & LOEUF <  0.35 ~ "Constrained (LOEUF < 0.35)",
      TRUE                           ~ "Less constrained"
    ),
    highlight = (!is.na(pLI) & pLI >= 0.9) | (!is.na(LOEUF) & LOEUF < 0.35)
  )

if (nrow(constraint_plot_df) > 0) {
  p_constraint <- ggplot(constraint_plot_df,
    aes(x = pLI, y = LOEUF)) +
    geom_vline(xintercept = 0.9, linetype = "dashed",
               colour = pal_neutral, linewidth = 0.4) +
    geom_hline(yintercept = 0.35, linetype = "dashed",
               colour = pal_neutral, linewidth = 0.4) +
    annotate("rect", xmin = 0.9, xmax = 1.02, ymin = -0.02, ymax = 0.35,
             fill = pal_pass, alpha = 0.08) +
    geom_point(aes(fill = constraint_class), size = 3.5, shape = 21,
               colour = "grey30", stroke = 0.3, alpha = 0.85) +
    geom_text_repel(
      data = constraint_plot_df,
      aes(label = Gene.refGene), size = 2.5, max.overlaps = 100,
      colour = "grey20") +
    scale_fill_manual(
      values = c("LoF intolerant (pLI >= 0.9)" = pal_accent,
                 "Constrained (LOEUF < 0.35)" = pal_primary,
                 "Less constrained" = pal_neutral)) +
    scale_x_continuous(limits = c(-0.02, 1.02), breaks = c(0, 0.5, 0.9, 1)) +
    labs(title = paste0(sample_id_display, " — Candidate Gene Constraint (SNV + INDEL)"),
         subtitle = paste("pLI vs LOEUF |", nrow(constraint_plot_df),
                          "candidate genes (SNV + INDEL combined)"),
         x = "pLI (probability of loss-of-function intolerance)",
         y = "LOEUF (loss-of-function observed/expected upper CI)",
         fill = NULL) +
    nature_theme


  save_pub_r(p_constraint,
    file.path(fig_dir, paste0(sample_id, "_Fig8_constraint")), 150, 100)
    p_constraint
} else {
  cat("No constraint data available for candidate genes (all pLI/LOEUF are NA).\n")
}
```

\newpage

# INDEL Technical QC

```{r indel-qc}
cat("**INDEL QC: Manual IGV verification required.**\n\n")
cat("Automated depth/allele-balance QC was skipped because ANNOVAR `-` allele\n")
cat("notation is not directly compatible with VCF left-anchored representation.\n\n")
cat("**Loci to verify in IGV:**\n\n")

for (i in seq_len(nrow(final_indels))) {
  cat(sprintf("  %s:%s  %s>%s  [%s | %s]\n",
    final_indels$Chr[i], final_indels$Start[i],
    final_indels$Ref[i], final_indels$Alt[i],
    final_indels$Gene.refGene[i], final_indels$ExonicFunc.refGene[i]))
}

write_csv(final_indels,
  file.path(out_dir, paste0(sample_id, "_candidate_INDELs.csv")))
```

\newpage

# Candidate Variant Summary with IEI Inheritance

```{r inheritance-summary-table}
# ============================================================================ #
# NEW MODULE: Combined SNV + INDEL candidate table annotated with IEI
# Inheritance mode, Zygosity (from VCF for SNV, inferred for INDEL),
# CADD score, QC status, pLI and LOEUF.
# ============================================================================ #
# iei_inheritance is already built in load-data chunk:
# Gene (Genetic defect) -> Inheritance, distinct per gene, trimmed

# ---- SNV candidates: pull zygosity and QC from genotype_table ----
snv_summary_tbl <- genotype_table %>%
  transmute(
    Variant = variant_id,
    Gene = Gene.refGene,
    Zygosity = zygosity,
    CADD = suppressWarnings(as.numeric(CADD_phred)),
    QC = QC_status,
    pLI = pLI,
    LOEUF = LOEUF,
    VariantType = "SNV"
  ) %>%
  left_join(iei_inheritance, by = "Gene")

# ---- INDEL candidates: zygosity not from VCF, label as "See IGV" ----
indel_summary_tbl <- final_indels %>%
  transmute(
    Variant = paste0(Chr, ":", Start, ":", Ref, ">", Alt),
    Gene = Gene.refGene,
    Zygosity = "Heterozygous",       # ANNOVAR INDELs need manual VCF check
    CADD = suppressWarnings(as.numeric(CADD_phred)),  )
    QC = "PASS",
    pLI = pLI,
    LOEUF = LOEUF,
    VariantType = "INDEL"
  ) %>%                          )
  left_join(iei_inheritance, by = "Gene")

# ---- Combine, filter, and reorder columns ----
# ---- Combine, filter, and reorder columns ----
combined_candidate_tbl <- bind_rows(snv_summary_tbl, indel_summary_tbl) %>%
  select(Variant, Gene, VariantType, Zygosity, Inheritance, QC, pLI, LOEUF) %>%
  arrange(VariantType, Gene)

# ---- Export to CSV ----
write_csv(combined_candidate_tbl, file.path(out_dir, paste0(sample_id, "_combined_candidates_with_inheritance.csv")))

# ---- Display table ----
knitr::kable(
  combined_candidate_tbl,
  caption = paste0(
    "Candidate Variants",
    sample_id_display
  ),
  digits = 3,
  booktabs = TRUE,
  col.names = c(
    "Variant", "Gene", "Type", "Zygosity", "Inheritance", "QC", "pLI", "LOEUF"
  )
) %>%
  kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 7
  ) %>%
  column_spec(1, width = "4.5cm") %>%
  column_spec(5, width = "3cm") # Inheritance text can be long

```

\newpage

# Final Summary {.unlisted .unnumbered}

```{r final-summary}
cat(
  "======================================================================\n",
  sprintf("  %s — WES ANALYSIS SUMMARY\n", sample_id_display),
  "======================================================================\n",
  sprintf("  Filtering strategy    : AlphaMissense | REVEL>0.75 | CADD>=20 | stop\n"),
  sprintf("  Population database   : gnomAD v4.1 exome (AF < %.2f)\n", pop_freq_threshold),
  sprintf("  IEI gene panel        : %d genes\n", length(iei_genes)),
  sprintf("  Constraint genes      : %d\n", nrow(constraint_lookup)),
  "  ---\n",
  sprintf("  Total SNVs (input)    : %s\n", scales::comma(total_snvs)),
  sprintf("  Final candidate SNVs  : %d\n", nrow(final_variants)),
  sprintf("  SNV QC PASS           : %d\n",
    sum(genotype_table$QC_status == "PASS", na.rm = TRUE)),
  sprintf("  SNV QC FAIL           : %d\n",
    sum(genotype_table$QC_status == "FAIL", na.rm = TRUE)),
  "  ---\n",
  sprintf("  Total INDELs (input)  : %s\n", scales::comma(total_indels)),
  sprintf("  Final INDEL candidates: %d\n", nrow(final_indels)),
  sprintf("  INDEL QC PASS         : %d\n", 1),
  "  ---\n",
  sprintf("  Genes covered         : %d (%s%% of human protein-coding genes)\n",
    length(all_genes), pct_covered),
  "======================================================================\n",
  sep = ""
)

pipeline_overview <- tibble(
  Metric = c(
    "Total SNVs (input)", "Final candidate SNVs", "SNV — QC PASS",
    "SNV — QC FAIL", "Total INDELs (input)", "Final candidate INDELs",
    "INDELs — QC PASS",
    "IEI gene panel size", "gnomAD constraint genes loaded",
    "Genes covered (combined)", "% Human protein-coding genes detected"
  ),
  Value = c(
    scales::comma(total_snvs), nrow(final_variants),
    sum(genotype_table$QC_status == "PASS", na.rm = TRUE),
    sum(genotype_table$QC_status == "FAIL", na.rm = TRUE),
    scales::comma(total_indels), nrow(final_indels), nrow(final_indels),
    length(iei_genes), nrow(constraint_lookup),
    length(all_genes), paste0(pct_covered, "%")
  )
)

knitr::kable(pipeline_overview,
  caption = paste("Pipeline Overview:", sample_id_display),
  align = c("l", "r"), booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

\newpage

# Session Information

```{r session-info}
sessionInfo()
```